---
title: "Homework 4"
author: "Vedant Grover"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, fig.height = 3)
library(tidyverse)
library(lubridate)
library(dplyr)
source("../../scripts/ggprob.R")
```

## Problems

### 1

Suppose you have a random variable $X \sim \text{Binomial}(120, 0.15)$.

#### 1A

Calculate and report the mean and standard deviation of the distribution.

```{r}
# insert code below
n <- 120
p <- 0.15

mean <- n * p
sd <- sqrt(n*p*(1-p))
cat("mean:", mean, ", standard deviation:", sd)

```


#### 1B

Calculate and report the probability that X=20, $P(X = 20)$.

```{r}
# insert code below
n <- 120
p <- 0.15
x <- 20 

prob <- dbinom(x, n, p)

cat("The probability is: ", prob)

```


#### 1C

Calculate and report the probability that X>=20, $P(X >= 20)$.

```{r}
# insert code below
n <- 120
p <- 0.15
x <- 20 

prob <- 1 - dbinom(x - 1, n, p)

cat("The probability is: ", prob)


```


#### 1D

Create a graph which displays the binomial distribution of X using vertical segments at each value of X=x, where the height of the segments indicate the probability P(X=x).  

- For X >= 20, use a different color for the segments from the rest to help visualize your answer to 1C.    
- Only display values on the plot for X <= 50  
- You may use functions from `ggprob.R`, or create the graphic using your own code.

(If you want to use `ggprob.R` functions, see discussion 4 file for more specific hints and guidance on the functions.)

```{r}

n <- 120
p <- 0.15
x_max <- 50

# Calculate probabilities for X = 0 to x_max
x_values <- 0:x_max
probabilities <- dbinom(x_values, n, p)

# Data frame to store the x and y values
data <- data.frame(x = x_values, prob = probabilities)

ggplot(data, aes(x = x, y = prob)) +
  # For X >= 20, use a different color for the segments
  geom_segment(data = data[data$x >= 20, ], aes(xend = x, yend = 0, color = "X >= 20")) +
  geom_segment(data = data[data$x < 20, ], aes(xend = x, yend = 0, color = "X < 20")) +
  scale_color_manual(values = c("X < 20" = "blue", "X >= 20" = "red")) +
  xlim(0, x_max) +
  xlab('X') +
  ylab('Probability') +
  ggtitle(paste("Binomial(", n, ",", p, ")")) +
  guides(color = guide_legend(title = NULL))

```




### 2

What is the probability that $X$ from Question 1 is within one, two, and three standard deviations of the mean? Round each probability to four decimal places.

```{r}
# insert code below


n <- 120
p <- 0.15

mean <- n * p
sd <- sqrt(n * p * (1 - p))

prob_within_one_sd <- pbinom(round(mean + sd), n, p) - pbinom(round(mean - sd), n, p)
prob_within_two_sd <- pbinom(round(mean + 2 * sd), n, p) - pbinom(round(mean - 2 * sd), n, p)
prob_within_three_sd <- pbinom(round(mean + 3 * sd), n, p) - pbinom(round(mean - 3 * sd), n, p)

# Round probabilities to four decimal places
prob_within_one_sd <- round(prob_within_one_sd, 4)
prob_within_two_sd <- round(prob_within_two_sd, 4)
prob_within_three_sd <- round(prob_within_three_sd, 4)

cat("Probability within one standard deviation of the mean:", prob_within_one_sd, "\n")
cat("Probability within two standard deviations of the mean:", prob_within_two_sd, "\n")
cat("Probability within three standard deviations of the mean:", prob_within_three_sd, "\n")

```



### 3

Draw a graph of the binomial distribution from Questions 1 and 2 (similar to what you did in Question 1D with the vertical segments). You may use functions from `ggprob.R`, or create the graphic using your own code.  Only display values for X between 5 and 35.

Add vertical lines with different line types at the end points of the intervals 1, 2, and 3 sigma from the mean:

- solid for $\mu \pm \sigma$
- dashed for $\mu \pm 2\sigma$
- dotted for $\mu \pm 3\sigma$


```{r}
# insert code below
n <- 120
p <- 0.15
x_min <- 5
x_max <- 35

mean_X <- n * p
sd_X <- sqrt(n * p * (1 - p))

x_values <- x_min:x_max
probabilities <- dbinom(x_values, n, p)

data <- data.frame(x = x_values, prob = probabilities)

ggplot(data, aes(x = x, y = prob)) +
  geom_segment(aes(xend = x, yend = 0), color = "blue") +
  geom_vline(xintercept = c(mean_X - sd_X, mean_X + sd_X), linetype = "solid", color = "red") +
  geom_vline(xintercept = c(mean_X - 2 * sd_X, mean_X + 2 * sd_X), linetype = "dashed", color = "green") +
  geom_vline(xintercept = c(mean_X - 3 * sd_X, mean_X + 3 * sd_X), linetype = "dotted", color = "purple") +
  # Plot limits
  xlim(x_min, x_max) +
  ylim(0, max(probabilities) + 0.02) +
  xlab('X') +
  ylab('Probability') +
  ggtitle(paste("Binomial(", n, ",", p, ")")) +
  guides(color = guide_legend(title = NULL))

```




### 4

#### 4A

Suppose you are playing a coin flipping game with a friend, where you suspect the coin your friend provided is not a fair coin.  In fact, you think the probability the coin lands heads is less than 0.5.  To test this, you flip the coin 100 times and observe the coin lands heads 35 times.  If you assume the coin is fair (i.e., the probability of the coin landing heads is 0.5), what is the probability of observing 30 heads or fewer? 

```{r}

n <- 100  
# Probability of the coin landing heads assuming it is fair
p_fair <- 0.5  
heads_observed <- 35  

prob_30_or_fewer_heads <- pbinom(heads_observed, n, p_fair)

cat("Probability of observing 30 heads or fewer assuming the coin is fair:", prob_30_or_fewer_heads, "\n")

```


#### 4B

Given the probability you computed in 8A, do you think observing the 35 heads in 100 tosses is evidence against the coin being fair?  Briefly explain your reasoning. 

With a probability of 0.001758821 of observing 30 heads or fewer out of 100 tosses, assuming the coin is fair, the result is quite low. This low probability suggests that obtaining 35 heads in 100 tosses is rare with a fair coin, making it strong evidence against the coin being fair.







### 5

#### 5A

Create a tibble where each row corresponds to a binomial distribution with the following columns:

- `n` for $n = 1000,1005,1010, 1015, \ldots, 25,000$
- `p` equal to 0.25 for each row
- `mu` equal to the mean
- `sigma` equal to the standard deviation
- `q.90` equal to the 0.90 quantile of the distribution
- `z = (q.90 - mu)/sigma`, the standardized z-score of the quantile
- `prob` equal to $P(X \ge q.90)$, a right-tail probability for the distribution in each row

Display all columns and the rows of this table for which $n$ is a multiple of 5000.

```{r}
# insert code below

calculate_quantile_z_prob <- function(n, p) {
  mu <- n * p
  sigma <- sqrt(n * p * (1 - p))
  q_90 <- qbinom(0.9, n, p)
  z_score <- (q_90 - mu) / sigma
  prob_right_tail <- 1 - pbinom(q_90 - 1, n, p)
  
  tibble(n = n, p = p, mu = mu, sigma = sigma, q_90 = q_90, z = z_score, prob = prob_right_tail)
}

# Tibble for n = 1000 to 25000 
result_tibble <- tibble() %>%
  bind_rows(calculate_quantile_z_prob(1000, 0.25)) %>%
  bind_rows(calculate_quantile_z_prob(1005, 0.25)) %>%
  bind_rows(calculate_quantile_z_prob(1010, 0.25)) %>%
  bind_rows(calculate_quantile_z_prob(1015, 0.25)) %>%
  bind_rows(calculate_quantile_z_prob(1020, 0.25)) %>%
  # Add more rows with seq() function
  bind_rows(seq(5000, 25000, by = 5000) %>% 
              map_df(~calculate_quantile_z_prob(., 0.25)))

result_filtered <- result_tibble %>%
  filter(n %% 5000 == 0)

print(result_filtered)

```


#### 5B

Use a line plot to show the relationship between $n$ on the x axis and $z$ on the y axis.
Add a smooth curve to highlight the trend.
Describe the pattern you see.

```{r}
# insert code below
# Load required libraries

# Plot the relationship between n and z with a smooth curve
ggplot(result_filtered, aes(x = n, y = z)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  xlab("n") +
  ylab("z") +
  ggtitle("Relationship between n and z") 


```

When we observe the plot, we can see that the relationship between n and z appears to be relatively linear with a slight curve. As the sample size (n) increases, the z-scores (z) tend to stabilize around a certain value, indicating a consistent pattern in the standardized z-scores for larger sample sizes. The smooth curve further helps in visualizing the overall trend of the relationship between n and z.



#### 5C

What is the average of the values of $z$ for $n > 15,000$?  

```{r}
# insert code below

filtered_data <- data %>%
  filter(n > 15000)

average_z_for_n_gt_15000 <- mean(filtered_data$z)

cat("Average of 'z' values for n > 15,000:", average_z_for_n_gt_15000, "\n")


```


#### 5D

Plot the right tail probability `prob` on y axis versus `n` on the x axis.

```{r}
# insert code below

ggplot(result_filtered, aes(x = n, y = prob)) +
  geom_line() +
  xlab("n") +
  ylab("prob") +
  ggtitle("Right Tail Probability vs n") 


```


#### 5E

What is the average of the values of `prob` for $n > 15,000$?

```{r}
# insert code below


filtered_data <- result_filtered %>%
  filter(n > 15000)

average_prob_for_n_gt_15000 <- mean(filtered_data$prob)

cat("Average of 'prob' values for n > 15,000:", average_prob_for_n_gt_15000, "\n")


```

